---
title: "NHANES blood pressure — Linear Regression"
author: "T.V.Q (James)"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: tango
    css: style.css
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 9,
  fig.height = 5,
  dpi = 150
)
```

## Introduction
Linear regression analysis is used to predict the value of a variable (dependent variable) based on the value of 
another variable(s) (dependent variable(s)) by fitting a linear equation to observed data. This form of analysis 
estimates the coefficients of the linear equation, involving one or more independent variables. <br>

Systolic blood pressure a clinically important outcome it is the strongest predictor of cardiovascular disease (CVD),
mortality, and target organ damage, especially in adults over 50. Elevated SBP represents the force of blood 
against arterial walls, and its control is critical for reducing stroke, heart failure, and renal failure risk <br>

## Data
The dataset used in this report is the NHANES dataset from `NHANES` package in `R`. It is a subset of the data of a yearly 
survey conducted by the US National Center for Health Statistics (NCHS) since 1999. The outcome of interest is the Systolic
blood pressure and its linear model is fitted against Age, BMI and gender. The reference level for used for gender is female. <br>

The dataset is filtered for adult population (age >= 18) and complete-case analysis of Systolic blood pressure, Age, BMI and gender
because of its simplicity and there is no assumption of mechanism of missing data.

```{r}
# Load the library
library(NHANES)     # For NHANES dataset
library(dplyr)      # For data manipulation
library(ggplot2)    # For graphs
library(patchwork)  # To combine multiple plots into a single layout
library(broom)      # For tidy() function
library(gtsummary)  # For publication-ready regression tables
library(knitr)      # For kable() function
```

```{r}
nhanes <- NHANES |>
  filter(Age >= 18) |>
  distinct() |>
  filter(!is.na(BPSysAve), !is.na(Age), !is.na(BMI), !is.na(Gender)) |>
  mutate(Gender = factor(Gender, levels = c("female", "male")))
```


## Exploratory Analysis
```{r, fig.cap = "Figure 1. Scatter plot of Systolic BP and Age — NHANES adults (over 18 years old)"}
# Scatter plot with linear fit and the LOESS curve
#   Variables: BPSysAve vs Age
ggplot(nhanes, aes(x = Age, y = BPSysAve)) +
  geom_point(alpha = 0.15, size = 0.6) +
  geom_smooth(method = "lm", color = "steelblue") +
  geom_smooth(method = "loess", color = "firebrick", linetype = "dashed") +
  labs(title = "BP vs Age", x = "Age (years)", y = "Systolic BP (mmHg)") +
  theme_minimal()
```


```{r, fig.cap = "Figure 2. Scatter plot of Systolic BP and BMI — NHANES adults (over 18 years old)"}
# Scatter plot with linear fit and the LOESS curve
#   Variables: BPSysAve vs BMI
ggplot(nhanes, aes(x = BMI, y = BPSysAve)) +
  geom_point(alpha = 0.15, size = 0.6) +
  geom_smooth(method = "lm", color = "steelblue") +
  geom_smooth(method = "loess", color = "firebrick", linetype = "dashed") +
  labs(title = "BP vs BMI", x = "BMI (kg/m²)", y = "Systolic BP (mmHg)") +
  theme_minimal()
```

```{r, fig.cap = "Figure 3. Box plot of Systolic BP by gender — NHANES adults (over 18 years old)"}
# Box plot
#   Variables: BPSysAve vs BMI
ggplot(nhanes, aes(x = Gender, y = BPSysAve, fill = Gender)) +
  geom_boxplot(alpha = 0.6, outlier.size = 0.5) +
  labs(title = "BP by Gender", x = NULL, y = "Systolic BP (mmHg)") +
  theme_minimal() + theme(legend.position = "none")
```


## Model Comparison

```{r}
# Build the models
#   Model 1: Age only
m1 <- lm(BPSysAve ~ Age, data = nhanes)

#   Model 2: Age + Gender
m2 <- lm(BPSysAve ~ Age + Gender, data = nhanes)

#   Model 3: Age + Gender + BMI (full model)
m3 <- lm(BPSysAve ~ Age + Gender + BMI, data = nhanes)
```

```{r}
# Report the summary table
bind_rows(
  glance(m1) |> mutate(model = "M1: Age"),
  glance(m2) |> mutate(model = "M2: Age + Gender"),
  glance(m3) |> mutate(model = "M3: Age + Gender + BMI")
) |>
  select(model, r.squared, adj.r.squared, AIC, BIC) |>
  mutate(across(where(is.numeric), ~ round(., 3)))  |>
  kable(caption = "Table 1. Summary of linear regression models of 
        Average Systolic BP against Age, BMI and gender — NHANES adults 
              (over 18 years old)",
    col.names = c("Model", "R²", "adjusted R²", "AIC", "BIC")
  )
```


The `R²` of model 1 (simple model), 2 and 3 (full model) are 0.187, 0.206, and 0.215, respectively,
which means 18.7%, 20.6%. and 21.5% of the variation of Systolic BP is explained by model 1, 2, and 3,
respectively.

The full model fits best because it has the higest `R²` and adjusted `R²` <br>

## Coefficient Interpretation
```{r, fig.cap = "Figure 4. Regression coefficients — predictors of systolic blood pressure — NHANES adults (over 18 years old)"}
# Round all numeric columns in smmary table of model 3 to 3 decimal places
m3_tidy <- tidy(m3, conf.int = TRUE) |>
  mutate(across(where(is.numeric) & !p.value, ~ round(., 3))) |>
  mutate(p.value = format.pval(p.value, digits = 3))

# Coefficient plot
m3_tidy |>
  filter(term != "(Intercept)") |>  # Remove the row of the intercept
  mutate(term = recode(term,
    "Age"        = "Age (per year)",
    "Gendermale" = "Gender: Male vs Female",
    "BMI"        = "BMI (per kg/m²)"
  )) |>
  ggplot(aes(x = estimate, y = term)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), orientation = "y",
                height = 0.2, linewidth = 0.8, color = "steelblue") +
  geom_point(size = 3, color = "steelblue") +
  geom_text(aes(label = paste0("β = ", estimate, "\n95% CI [",
      conf.low, ", ", conf.high, "]"
    )), hjust = -0.1, size = 3, color = "grey30"
  ) +
  labs(
    title   = "Regression coefficients — predictors of systolic blood pressure",
    subtitle = "Model 3: BPSysAve ~ Age + Gender + BMI | NHANES adults",
    x        = "Coefficient estimate (mmHg)",
    y        = NULL
  ) +
  theme_minimal() +
  xlim(min(m3_tidy$conf.low[m3_tidy$term != "(Intercept)"], na.rm = TRUE) - 1,
       max(m3_tidy$conf.high[m3_tidy$term != "(Intercept)"], na.rm = TRUE) + 3)
```
* Holding all other variables constant, a one-unit increase in Age is associated with a 0.421 mmHg increase in systolic BP (95% CI: [0.398, 0.444], p < 2e-16)
* Holding all other variables constant, a one-unit increase in BMI is associated with a 0.253 mmHg increase in systolic BP (95% CI: [0.192, 0.313], p = 3.09e-16)
* Males have, on average, 4.81 mmHg higher systolic BP than females of the same age and BMI (95% CI: [4.00 , 5.63], p < 2e-16)

All variables (Age, BMI, and gender) are statistically significant

## Assumption Checks
```{r, fig.cap = "Figure 5. Assumption plots for linear regression of Systolic blood pressure — NHANES adults (over 18 years old)"}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(m3)
```
* Residuals vs Fitted plot shows that points scattered quite randomly around the horizontal zero line, no funnel shape, no curve <br>
  => Linearity and homoscedasticity assumptions are fulfilled
* Normal Q-Q plot shows that points follow the diagonal line in the middle. However, points largely break away from the line at the top right
  it might indicate that the normal model does not accurately describe the heavy tail (right-skewed) of data <br>
  => Normality assumption might not be fulfilled <br>
  => Log - Transformation of outcome can be used 
* Scale-Location plot shows that red line is roughly horizontal across the plot and points spread quite evenly <br>
  => Homoscedasticity assumption is fulfilled
* Residuals vs Leverage plot shows that there is no points outside Cook's distance lines <br>
 => There are no influential observations


## Predicted Effects
```{r, fig.cap = "Figure 6. Predicted effects plots of Systolic blood pressure — NHANES adults (over 18 years old)"}
# Create a prediction grid for Age (holding BMI at median, Gender = female)
age_grid <- data.frame(
  Age    = seq(min(nhanes$Age), max(nhanes$Age), length.out = 100),
  BMI    = median(nhanes$BMI),
  Gender = factor("female", levels = c("female", "male"))
)


age_pred <- predict(m3, newdata = age_grid, interval = "confidence") |>
  as.data.frame() |>
  bind_cols(age_grid)


p_partial_age <- ggplot(age_pred, aes(x = Age)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = "steelblue", alpha = 0.2) +
  geom_line(aes(y = fit), color = "steelblue", linewidth = 1) +
  labs(
    title    = "Predicted systolic BP by age",
    subtitle = "BMI held at median; Gender = Female",
    x        = "Age (years)", y = "Predicted systolic BP (mmHg)"
  ) +
  theme_minimal()

# Repeat for BMI (holding Age at mean, Gender = female)
bmi_grid <- data.frame(
  BMI    = seq(min(nhanes$BMI), max(nhanes$BMI), length.out = 100),
  Age    = mean(nhanes$Age),
  Gender = factor("female", levels = c("female", "male"))
)

bmi_pred <- predict(m3, newdata = bmi_grid, interval = "confidence") |>
  as.data.frame() |>
  bind_cols(bmi_grid)

p_partial_bmi <- ggplot(bmi_pred, aes(x = BMI)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = "coral", alpha = 0.2) +
  geom_line(aes(y = fit), color = "coral", linewidth = 1) +
  labs(
    title    = "Predicted systolic BP by BMI",
    subtitle = "Age held at mean; Gender = Female",
    x        = "BMI (kg/m²)", y = "Predicted systolic BP (mmHg)"
  ) +
  theme_minimal()

# Plot the figures
p_partial_age | p_partial_bmi
```
* When BMI is held at median and gender is female, as Age increases, the outcome of interest Systolic BP increases.
  The Prediction interval is close with the regression line indicating that there is a good precision with predicting Systolic BP from Age

* When Age is held at mean and gender is female, as BMI increases, the outcome of interest Systolic BP increases.
  The Prediction interval is wider than that of the plot of Age so there is more uncertainty in the prediction of Systolic BP from BMI.
  Especially, the Prediction interval is wider at the top right of the plot indicating that higher BMI tends not to be good predictor of Systolic BP

## Publication Table
```{r}
tbl_regression(m3,
  label = list(
    Age        ~ "Age (years)",
    Gender     ~ "Gender",
    BMI        ~ "BMI (kg/m²)"
  )
) |>
  add_glance_source_note(
    label = list(r.squared ~ "R²", adj.r.squared ~ "Adjusted R²"),
    include = c(r.squared, adj.r.squared, nobs)
  ) |>
  bold_p(t = 0.05) |>
  bold_labels()
```
21.5% of the variation of Systolic BP is explained by lineal model including Age, BMI and gender


## Limitations
There are several Limitations with this analysis:
* The linearity assumptions are not fulfilled totally. The Normality assumption might not be fulfilled as 
  the normal model does not accurately describe the heavy tail (right-skewed) of data

* The linear model only assess 3 variables without considering interaction or other confounders.

* The dataset is a cross-sectional survey, it cannot hold any conclusion for the trend of the effects over time.


## Bonus: Test whether the relationship between Age and BP is non-linear
```{r}
# Create a column of the quadratic term of Age
nhanes <- nhanes |> mutate(Age2 = Age^2)

# Model 4: Age + Gender + BMI + Age²
m4 <- lm(BPSysAve ~ Age + Age2 + Gender + BMI, data = nhanes)

# Compare M3 vs M4 with a likelihood ratio test
anova(m3, m4)
```
With p-value of 3.283e-16, adding the quadratic term of Age contribute significantly to the model including Age, BMI and gender.
Therefore, relationship between Age and Systolic BP is non-linear


## Session Info
```{r session-info}
sessionInfo()
```