---
title: "NHANES Health Indicators — Descriptive Analysis"
author: "T.V.Q (James)"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 9,
  fig.height = 5,
  dpi = 150
)
```

## Introduction
NHANES dataset is a subset of the data of a yearly survey conducted by the US National Center for Health Statistics (NCHS) since 1999. It contains 10,000 rows of data collected in 2009-2010 and 2011-2012 with 75 variables. <br>
This report is to present the descriptive statistics of key health indicators: BMI, Age, Systolic and Diastolic blood pressure, and Total Cholesterol of the adult population in the NHANES dataset. <br>
The adult population is defined as having age equal or over 18 years old.

## Data preparation
```{r}
# Load the library for NHANES dataset
library(NHANES)

# Load the library for data manipulation
library(dplyr)

# Filter adult participants (age ≥ 18) and remove duplicate rows
nhanes_adult <- NHANES |>
  filter(Age >= 18) |>
  distinct()

# Number of remaining observartions
nrow(nhanes_adult)
```
After filtering for adult participants and removing duplicate observartions, there are 5679 observations in the dataset. <br>
Duplicate observartions are defined as having with identical values across all variables.
These observartions should be removed to not use the information of one observation more than once. <br>

## Descriptive statistics
```{r}
# Load the library for function skewness()
## to calculate the skewness of a distribution
library(moments)

# Load the library for bind_rows() function
## to bind data row into a table
library(dplyr)

# Load the library for kable() function
## to report a data frame as a formatted table
library(knitr)

# Define a custom function for descriptive statistics
## The input of the function should be vector
describe_numeric <- function(x, varname = "variable") {
  x_clean <- x[!is.na(x)]
  q1  <- quantile(x_clean, 0.25)
  q3  <- quantile(x_clean, 0.75)
  iqr <- q3 - q1

  data.frame(
    variable   = varname,
    n          = length(x_clean),
    n_missing  = sum(is.na(x)),
    pct_missing = round(mean(is.na(x)) * 100, 1),
    mean       = round(mean(x_clean), 2),
    median     = round(median(x_clean), 2),
    sd         = round(sd(x_clean), 2),
    p25        = round(q1, 2),
    p75        = round(q3, 2),
    iqr        = round(iqr, 2),
    min        = round(min(x_clean), 2),
    max        = round(max(x_clean), 2),
    skewness   = round(skewness(x_clean), 3),
    n_outliers = sum(x_clean < (q1 - 1.5 * iqr) | x_clean > (q3 + 1.5 * iqr))
  )
}

# Create a summary table
summary_table <- bind_rows(
  describe_numeric(nhanes_adult$BMI,      "BMI"),
  describe_numeric(nhanes_adult$Age,      "Age"),
  describe_numeric(nhanes_adult$BPSysAve, "Systolic BP"),
  describe_numeric(nhanes_adult$BPDiaAve, "Diastolic BP"),
  describe_numeric(nhanes_adult$TotChol,  "Total Cholesterol")
)

# Report the summary table
kable(summary_table,
      digits  = 2,
      caption = "Table 1. Descriptive statistics for 
                  key health indicators — NHANES adults (n ≥ 18)",
      col.names = c("Variable", "N", "Missing", "% Missing",
                    "Mean", "Median", "SD", "P25", "P75",
                    "IQR", "Min", "Max", "Skewness", "N Outliers"))
```
* The BMI is the most skewed variable with the skewness of 1.333. The positive skewness indicates that the distribution of BMI is highly right-skewed where the mean is largely greater than the median. The distribution of BMI might not be normal distribution.
* The BMI also has the most outliers by the 1.5 × IQR rule with 172 outliers.
* The Age has the most symmetrical shape among the examined indicators with the skewness of 0.24. All other variables are highly skewed
* The Age also has the least number of outliers, which is 0

## Distributions by Gender
```{r, fig.cap = "Figure 1. Distribution of key health indicators by gender — NHANES"}
# Load the library for function pivot_longer()
## to reshape the dataset to long format
library(tidyr)

# Load the library for creating graph
library(ggplot2)

# Reshape the dataset to long format using pivot_longer()
nhanes_long <- nhanes_adult |>
  select(Gender, BMI, Age, BPSysAve, BPDiaAve, TotChol) |>
  pivot_longer(cols = -Gender, names_to = "variable", values_to = "value") |>
  filter(!is.na(value), !is.na(Gender))

# Plot the boxplots
## use facet_wrap() to split a plot into multiple plots
## based on a categorical variable
ggplot(nhanes_long, aes(x = Gender, y = value, fill = Gender)) +
  geom_boxplot(outlier.size = 0.8, outlier.alpha = 0.4, outlier.color = "red") +
  facet_wrap(~variable, scales = "free_y") +
  labs(title = "Distribution of key health indicators by gender — NHANES",
       x = NULL, y = "Value") +
  theme_minimal() +
  theme(legend.position = "bottom")
```
There are no obvious differences of the distribution of all 5 key health indicators across gender. 
The distributions between 2 genders are mostly overlap where the medians and the interquartile ranges are mostly similar with each indicators.

## BMI and Age
```{r, fig.cap = "Figure 2. BMI vs age by gender — NHANES adults"}
# Filter the dataset without missing value in BMI, Age, and Gender columns
nhanes_bmi <- nhanes_adult |>
  filter(!is.na(BMI), !is.na(Age), !is.na(Gender))

# Plot the scatter plot
ggplot(nhanes_bmi, aes(x = Age, y = BMI, color = Gender)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 1.2) +
  ## method = "loess" means that smoothing line uses LOESS method
  ### Locally Estimated Scatterplot Smoothing
  ### se = TRUE means displaying the confidence band
  labs(title = "BMI vs age by gender — NHANES adults",
       x = "Age (years)", y = "BMI (kg/m²)") +
  theme_minimal()
```
* The LOESS curve plateaus with age. Therefore, when age increases, the BMI is not affected
* The trend is similar for both gender. The LOESS curves for both gender do not split apart, suggesting that BMI is not affected by gender
* The width of the confidence band for both genders is narrow, indicate high precision. However, it can be affected by the large sample size (over 5,000)

## Missing Data Summary
```{r}
# Filter variables with missing observation from summary_table
summary_missing <- summary_table |>
  select(variable, n_missing, pct_missing) |>
  filter(n_missing > 0)

# Report the summary table
kable(summary_missing,
      digits  = 2,
      caption = "Table 2. Summary of missing data for 
                  key health indicators — NHANES adults (n ≥ 18)",
      col.names = c("Variable", "Missing", "% Missing"))
```
* The variable has the most missing data is the Total Cholesterol `TotChol` with 320 missing observations, accounting for 5.6%
* Only Age has no missing data

## Key Takeaways
There are 5 key indicators examined with the dult population in the NHANES dataset:  BMI, Age, Systolic and Diastolic blood pressure, and Total Cholesterol <br>

* The data of adult population is quite complete.
Among the five health indicators, there are no indicators with more than 10% of missing data. 
Therefore, reliable conclusions could be drawn from this population without serious bias from missing data.
* All of the examined indicators are skewed with the Age is the least skewed indicators. 
Therefore, one should be caution when using normal assumption.
* The distributions of these indicators are not affected by gender. The distributions between 2 genders are mostly overlap
* The LOESS curve between Age and BMI plateaus when Age increases, indicating that BMI is not affected by Age.
The trend is similar in both genders.

## Session Info
```{r session-info}
sessionInfo()
```